Red Wine Quality Analysis by Shantanu Kulkarni
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
```

```{r}
# Load the Data
wineQualityReds <- read.csv("~/wineQualityReds.csv")
str(wineQualityReds)

```
Removing column 'X' from data frame, we run basic summary statistics
```{r}
wineQualityReds$X <- NULL
summary(wineQualityReds)

```
After running the above commands, here are some simple observations:

* There are 1599 observations in the dataset with 12 variables.
* Column 'X' was an identifier which wasn't useful for analysis, so I removed it from the dataset.
* Quality variable is my focus of interest and it is an ordered, categorical variable. As per the information given     regarding the dataset, red wines are rated on a 0-10 scale, and our sample had a range of 3-8 with a mean of 5.6      and median of 6
* All other variables are continuous in nature.

Since we are interested in the categorical variable Quality ,it makes sense to convert it into a factor.
```{r}
wineQualityReds$quality <- factor(wineQualityReds$quality, ordered = T)
str(wineQualityReds$quality)
table(wineQualityReds$quality)
```

The above command shows us that majority of the wines have been rated between 5 and 7.

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
grid.arrange(qplot(wineQualityReds$fixed.acidity),
             qplot(wineQualityReds$volatile.acidity),
             qplot(wineQualityReds$citric.acid),
             qplot(wineQualityReds$residual.sugar),
             qplot(wineQualityReds$chlorides),
             qplot(wineQualityReds$free.sulfur.dioxide),
             qplot(wineQualityReds$total.sulfur.dioxide),
             qplot(wineQualityReds$density),
             qplot(wineQualityReds$pH),
             qplot(wineQualityReds$sulphates),
             qplot(wineQualityReds$alcohol),
             qplot(wineQualityReds$quality),
             ncol = 4)
```

The histogram reveals following observations:

* Density and pH are normally distributed with a few outliers.
* Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol seem to be long-tailed.

Let's proceed with creating boxplots.

```{r,echo=FALSE}
grid.arrange(qplot(x = 'Fixed Acidity',
                   y = wineQualityReds$fixed.acidity,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Volatile Acidity',
                   y = wineQualityReds$volatile.acidity,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Citric Acid',
                   y = wineQualityReds$citric.acid,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Residual Sugar',
                   y = wineQualityReds$residual.sugar,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Chlorides',
                   y = wineQualityReds$chlorides,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Free Sulfur Dioxide',
                   y = wineQualityReds$free.sulfur.dioxide,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Total Sulfur Dioxide',
                   y = wineQualityReds$total.sulfur.dioxide,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Density',
                   y = wineQualityReds$density,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'pH',
                   y = wineQualityReds$pH,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Sulphates',
                   y = wineQualityReds$sulphates,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Alcohol',
                   y = wineQualityReds$alcohol,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             qplot(x = 'Quality',
                   y = wineQualityReds$quality,
                   data = wineQualityReds,
                   geom = 'boxplot',
                   xlab = '', ylab = ''),
             ncol = 4)
```

Boxplots confirm most observations seen in histograms.

# Univariate Analysis
Since we are majorly interested in Quality variable, it makes sense to create another categorical variable called wine rating that will contain 'bad' (rated from 0-4),'average'(rated from 5-7) and'good'(rated greater 7 or greater) wines.

```{r, echo=FALSE}
wineQualityReds$WineRating <- ifelse(wineQualityReds$quality >= 7, 'good', ifelse(wineQualityReds$quality <= 4, 'bad', 'average'))

wineQualityReds$WineRating <- ordered(wineQualityReds$WineRating, levels = c('bad', 'average', 'good'))

summary(wineQualityReds$WineRating)

ggplot(aes(wineQualityReds$WineRating),
       data = wineQualityReds) +
  geom_bar()
```

In line with histogram, we can see most wines are average in rating.

### What is the structure of your dataset?

* There are 1599 observations in the dataset with 12 variables.

* The 12 features and their short description is as follows: 

* Fixed acidity: most acids involved with wine or fixed or nonvolatile (do not    evaporate readily)
* Volatile acidity: the amount of acetic acid in wine
* Citric acid: mostly used for adding flavor
* Residual sugar: the amount of sugar remaining after fermentation stops and      measured in gms of sugar per liter.
* Chlorides: the amount of salt in the wine
* Free sulfur dioxide: the free form of SO2 exists in equilibrium between         molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial     growth and the oxidation of wine
* Total sulfur dioxide: amount of free and bound forms of S02
* Density: the density of water is close to that of water depending on the        percent alcohol and sugar content
* pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to   14 (very basic); most wines are between 3-4 on the pH scale
* Sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)     levels, wich acts as an antimicrobial and antioxidant
* Alcohol: the percent alcohol present in the  wine
* Quality: rating of the wine between 0-10 in the given sample.

### What is/are the main feature(s) of interest in your dataset?

* Quality of the wine is my main focus here. Apart from that, I will also take a   look at how other variables affect quality of the wine
* As per this article, sweetness, acidity, tannin, fruit and alcohol content are   the basic characteristics of a wine. While our dataset do not have all          features, I will look into all other features that may play an important role   in rating a wine. 

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
* As mentioned above, I look forward to exploring features like alcohol content,   acidity, pH, sugars and how they affect quality of the wine.
* The histogram reveals following observations:
  * Density and pH are normally distributed with a few outliers.
  * Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol are         long-tailed.

### Did you create any new variables from existing variables in the dataset?
* Since we are majorly interested in Quality variable, it makes sense to create   another categorical variable called wine rating that will contain 'bad' (rated   from 0-4),'average'(rated from 5-7) and'good'(rated greater 7 or greater)       wines.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section

Now let's create stacked histogram of different attributes for wine rating values. I have only created histograms for attributes that might impact quality of wine as per this article.


```{r}
#Stacked histogram of different attributes based on wine rating feature.

# Histogram of residual sugar based on wine rating
p1<-ggplot(wineQualityReds,aes(x=wineQualityReds$residual.sugar,fill=WineRating))+
  geom_histogram()

#Histogram of alcohol content for different wine rating values
p2<-ggplot(wineQualityReds,aes(x=wineQualityReds$alcohol,fill=WineRating))+
  geom_histogram()

#Histogram of pH content for different wine rating 
p3<-ggplot(wineQualityReds,aes(x=wineQualityReds$pH,fill=WineRating))+
  geom_histogram()

#Histogram of chloride content for different wine rating 
p4<-ggplot(wineQualityReds,aes(x=wineQualityReds$chlorides,fill=WineRating))+
  geom_histogram()

grid.arrange(p1,p2,p3,p4, ncol=2)
```

Unfortunately, the histograms do not reveal much information except the fact that good wines have higher amount of alcohol. Let's go ahead and find correlation among quality and other features.

```{r}
cor(x=wineQualityReds[,1:11], y=as.numeric(wineQualityReds$quality))

```

From the results, we see that alcohol,among all other,stands out in terms of positive linear relationship. Apart from that,citric acid and sulphates also show positive relationship. Volatile acidity stands out as a negatively correlated variable.Let's create some scatterplots to examine above relationships visually.

```{r, echo=FALSE}
p5<-ggplot(aes(x = wineQualityReds$quality, y = wineQualityReds$alcohol),
       data = wineQualityReds) +
  geom_smooth(method='lm', aes(group = 1))+
  geom_point(alpha = 1/3, position = position_jitter(h = 0)) + 
  xlab('Quality of wine') +
  ylab('Alcohol content')

p6<-ggplot(aes(x = wineQualityReds$quality, y = wineQualityReds$volatile.acidity),
       data = wineQualityReds) +
  geom_smooth(method='lm', aes(group = 1))+
  geom_point(alpha = 1/3, position = position_jitter(h = 0)) + 
  xlab('Quality of wine') +
  ylab('Volatile Acidity')

p7<-ggplot(aes(x = wineQualityReds$quality, y = wineQualityReds$sulphates),
       data = wineQualityReds) +
  geom_smooth(method='lm', aes(group = 1))+
  geom_point(alpha = 1/3, position = position_jitter(h = 0)) + 
  xlab('Quality of wine') +
  ylab('Sulphates')

p8<-ggplot(aes(x = wineQualityReds$quality, y = wineQualityReds$citric.acid),
       data = wineQualityReds) +
  geom_smooth(method='lm', aes(group = 1))+
  geom_point(alpha = 1/3, position = position_jitter(h = 0)) + 
  xlab('Quality of wine') +
  ylab('Citric Acid')

grid.arrange(p5,p6,p7,p8,ncol=2)
```


```{r}
p9<-ggplot(data=wineQualityReds, aes(x=as.factor(wineQualityReds$quality), y=wineQualityReds$alcohol)) +
    geom_boxplot()+
    xlab("Wine Quality") +
    ylab("Alcohol")

p10<-ggplot(data=wineQualityReds, aes(x=as.factor(wineQualityReds$quality), y=wineQualityReds$volatile.acidity)) +
    geom_boxplot()+
    xlab("Wine Quality") +
    ylab("Volatile acidity")

p11<-ggplot(data=wineQualityReds, aes(x=as.factor(wineQualityReds$quality), y=wineQualityReds$citric.acid)) +
    geom_boxplot()+
    xlab("Wine Quality") +
    ylab("Citric Acid")

grid.arrange(p9,p10,p11,ncol=3)

```

Scatterplots and boxplots confirm our previous finding that:

* Good wine generally has higher alcohol content,citric acidity and lower volatile acidity.This makes sense, because higher volatile acidity leads to an   unpleasant taste and lower alcohol level doesn't live up to the reputation of what is considered as a good wine.

```{r}
set.seed(1836)
View(wineQualityReds)
winered_subset <- wineQualityReds[,c(1:12)]
cor(winered_subset[, c(1:11)])
```

```{r}
p12<-ggplot(data = wineQualityReds, aes(x = citric.acid, y = fixed.acidity)) +
  geom_jitter(alpha=1/4, color = 'green') +
  geom_smooth(method='lm', color='red')

p13<-ggplot(data = wineQualityReds, aes(x = fixed.acidity, y = density)) +
  geom_jitter(alpha=1/3, color = 'red') +
  geom_smooth(method='lm', color='blue')


p14<-ggplot(data = wineQualityReds, aes(x = citric.acid, y = volatile.acidity)) +
  geom_jitter(alpha=1/3, color = 'blue') +
  geom_smooth(method='lm', color='red')

p15<-ggplot(data = wineQualityReds, aes(x = fixed.acidity, y = pH)) +
  geom_jitter(alpha=1/3, color = 'green') +
  geom_smooth(method='lm', color='blue')

grid.arrange(p12,p13,p14,p15,ncol=2)

```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

* Stacked histograms of variables with respect to quality did not reveal much information except     that alcohol strongly affects wine quality.
* Correlation test, scatterplots and boxplots showed strong positive correlation of alcohol with     quality and strong negative correlation between volatile acidity and quality.
* Thus leads to a general observation that a Good wine generally has higher alcohol content,higher   citric acidity and lower volatile acidity

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

* From the correlation matrix,following strong correlations for some of other features can be seen:

* Fixed acidity vs citric acid (0.67)
* Volatile acidity vs citric acid (-0.55)
* Fixed acidity vs density (0.67)
* Fixed acidity vs pH (-0.68)


### What was the strongest relationship you found?
* Citric acid and fixed acidity showed a strong positive correlation (68%) whereas pH and fixed      acidity showed a strong negative correlation(-68%)

# Multivariate Plots Section


```{r, echo=FALSE}
mvplot1<- ggplot(aes(x = citric.acid, y = fixed.acidity, color = quality), 
       data = wineQualityReds) + 
  geom_point() +
  facet_wrap(~WineRating) 

mvplot2<- ggplot(aes(x = citric.acid, y = volatile.acidity, color = quality), 
       data = wineQualityReds) + 
  geom_point() +
  facet_wrap(~WineRating) 

mvplot3<- ggplot(aes(x = fixed.acidity, y = density, color = quality), 
       data = wineQualityReds) + 
  geom_point() +
  facet_wrap(~WineRating) 

mvplot4<- ggplot(aes(x = fixed.acidity, y = pH, color = quality), 
       data = wineQualityReds) + 
  geom_point() +
  facet_wrap(~WineRating) 

grid.arrange(mvplot1,mvplot2,mvplot3,mvplot4,ncol=2)
```



```{r}

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

* For multivariate analysis, non-interest features that were highly corelated to each other were     visually examined and faceted by quality variable. 
* Good quality wines tend to have higher citric acid and lower volatile acidity levels.
* I expected pH to play a bigger role in determining wine quality but the sample doesn't seem to     indicate that.
* Alcohol, citric acid,volatile acidity,sulphates could be used to create a linear model to predict   quality of wines.

### Were there any interesting or surprising interactions between features?

* One of the surprising aspects of analysis is the pH did not really have a significant role to      play in determining wine quality. As such most good wines have pH between 3-4.
* Features that were highly correlated did not yield as much insight as expected, for example, d     density and fixed acidity when faceted by quality variable did not reveal any clear distinction    between different types of wines. Similarly, pH and fixed acidity when faceted by quality did not   provide adequate information.
* 
### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = citric.acid, y = fixed.acidity, color = WineRating),
       data = subset(wineQualityReds, WineRating != "average")) +
   coord_cartesian(xlim = c(quantile(wineQualityReds$citric.acid, 0),
                           quantile(wineQualityReds$citric.acid, 0.9))) +
  coord_cartesian(ylim = c(quantile(wineQualityReds$fixed.acidity, 0),
                           quantile(wineQualityReds$fixed.acidity, 0.9))) +
  geom_point() + 
  geom_vline(xintercept = mean(wineQualityReds$citric.acid),
             linetype_pal(3), color = 'red') +
  geom_hline(yintercept = mean(wineQualityReds$fixed.acidity),
             linetype_pal(3), color = 'dark blue') + 
  xlab('Citric Acid') +
  ylab('Fixed Acidity') + 
  ggtitle('Citric acid and fixed acidity across wines.')
```

### Description One
* Citric Acid and Fixed Acidity both play an important role in determining wine quality. It can be   clearly seen from the above plot that good quality wines generally have higher citric acid and     fixed acidity levels.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = alcohol, y = density, color = WineRating),
       data = subset(wineQualityReds, WineRating != "average")) +
   coord_cartesian(xlim = c(quantile(wineQualityReds$alcohol, 0),
                           quantile(wineQualityReds$alcohol, 0.9))) +
  coord_cartesian(ylim = c(quantile(wineQualityReds$density, 0),
                           quantile(wineQualityReds$density, 0.9))) +
  geom_point() + 
  geom_vline(xintercept = mean(wineQualityReds$alcohol),
             linetype_pal(3), color = 'red') +
  geom_hline(yintercept = mean(wineQualityReds$density),
             linetype_pal(3), color = 'dark blue') + 
  xlab('Alcohol') +
  ylab('Density') + 
  ggtitle('Alcohol and Density across wines.')  

```

### Description Two
* Above plot reveals that good wines generally have lower density and higher alcohol content

### Plot Three
```{r echo=FALSE, Plot_Three}
  ggplot(aes(x = alcohol, y = sulphates, color = WineRating),
       data = subset(wineQualityReds, WineRating != "average")) +
   coord_cartesian(xlim = c(quantile(wineQualityReds$alcohol, 0),
                           quantile(wineQualityReds$alcohol, 0.9))) +
  coord_cartesian(ylim = c(quantile(wineQualityReds$sulphates, 0),
                           quantile(wineQualityReds$sulphates, 0.9))) +
  geom_point() + 
  geom_vline(xintercept = mean(wineQualityReds$alcohol),
             linetype_pal(3), color = 'red') +
  geom_hline(yintercept = mean(wineQualityReds$sulphates),
             linetype_pal(3), color = 'dark blue') + 
  xlab('Alcohol') +
  ylab('Sulphates') + 
  ggtitle('Alcohol and Sulphates across wines.')  
```

### Description Three

* Above plot shows that good wines have a high amount of sulphates and alcohol.

------

# Reflection


